Energy Clustering

### 2017-10-24 energy_clustering/draft_v2/energy_clustering.tex
draft v2

\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
%\usepackage{nicefrac}       % compact symbols for 1/2, etc.
%\usepackage{microtype}      % microtypography
%\usepackage{mathtools}
%\usepackage{printlen}
%\usepackage{cite}
\usepackage[inline]{enumitem}
\usepackage[inline]{enumitem}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}
\usepackage{algorithmic}
\usepackage{algorithm}
%\usepackage[none]{hyphenat}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array,booktabs}
\usepackage{xcolor}
\usepackage{makecell}

@@

\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by the 
Newtonian gravitational potential from classical mechanics, and it provides 
a 
hypothesis test for equality of distributions. 
It was further generalized from Euclidean spaces to metric spaces of 
strong negative type, and more recently, a connection with reproducing 
kernel Hilbert spaces (RKHS) was established. 
Here we consider the problem of clustering data from an 
energy statistics theory perspective.
We provide a precise mathematical formulation 
yielding a quadratically constrained 
quadratic program (QCQP) in the associated RKHS. We show that this QCQP
is equivalent to kernel $k$-means optimization problem, once the kernel
is fixed.
Thus, our results imply a first principles derivation of kernel $k$-means 
from energy statistics.
However, energy statistics fixes a family of standard kernels.
Furthermore, we also consider a weighted version of energy statistics, 
making connection to graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative algorithm based 
on Hartigan's method, which in this case has the same computational cost 
as kernel $k$-means algorithm, based on Lloyd's method, but usually 
with better clustering quality. 
We provide carefully designed numerical experiments showing the superiority 
of the proposed method compared to kernel $k$-means, standard $k$-means, 
and Gaussian mixture models in a variety of settings.
\end{abstract}

% @gui: you didn't update the abstract?
% @jovo: I did, removed some stuff that you didn't like before, now
% removing the non-parametric word. You think it's not good enough?

@@

Energy statistics \cite{Szkely2013}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational potential in classical mechanics. 
It provides a non-parametric hypothesis test for equality of 
distributions which is achieved 
under minimum energy. When probability distributions are different the 
statistical potential energy diverges as sample size increases, while tends 
to a nondegenerate limit distribution when probability
distributions are equal. 
Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering \cite{RizzoClustering} 
by extending Ward's method of minimum variance. Moreover, an application of 
energy statistics to clustering was already proposed \cite{Kgroups}, 
which in part motivated this paper. We refer the reader to \cite{Szkely2013}, 
and references therein, for an overview of energy statistics theory and 
its applications.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
The notion of distance covariance was further generalized from Euclidean 
spaces to metric spaces of strong negative type \cite{Lyons}. Furthermore, 
the missing link between energy distance based tests and kernel 
based tests has 
been recently resolved \cite{Sejdinovic2013}, where a unifying framework
establishing an equivalence between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS), was established. 
This equivalence immediately relates energy statistics to
kernel methods, often used in machine learning, and form the basis 
of our approach.

@@

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen.
The well-known kernel $k$-means optimization problem is nothing but $k$-means 
in the feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd}
of grouping points that are closer to a cluster center, now
in feature space. 
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection with kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. Our results
immediately establish the connection with kernel methods, showing that this 
QCQP is equivalent to kernel $k$-means optimization problem. The 
equivalence between kernel $k$-means, spectral clustering, and graph 
partitioning problems is well-known \cite{Dhillon,Dhillon2}. We demonstrate 
how these relations arise from a weighted version of energy statistics.

Our main 
algorithmic contribution is to use Hartigan's method \cite{Hartigan} to 
find local solutions of the above mentioned QCQP, which is NP-hard in general.
Hartigan's method was also used in \cite{Kgroups}. More importantly, its 
advantages
over Lloyd's method was already demonstrated 
in some simple settings
\cite{Telgarsky,Slonin}, but apparently this method did not receive 
the deserved attention. To the 
best of our knowledge, Hartigan's method was not previously 
employed together with kernel methods. 
We provide a full kernel based Hartigan's algorithm for clustering,
where the kernel is fixed by energy statistics. 
We make clear the advantages of this proposal versus 
Lloyd's method, which kernel $k$-means is based upon and will also be used 
to solve our QCQP. We show that both algorithms  have the same
time complexity, but Hartigan's method in kernel spaces offer several
advantages. Furthermore, in the examples considered in this paper, it 
also provides superior performance compared to a spectral clustering
relaxation of the QCQP.

Our numerical results provide compelling evidence that 
Hartigan's method applied to energy statistics based clustering, 
or \emph{energy clustering} for short, is more accurate and robust than 
Lloyd's method. Our experiments also put in evidence the
flexibility of energy clustering, which provides a family of default kernels, 
showing that it is able to perform accurately on data coming from 
very different distributions, contrary to $k$-means and GMM for instance.
More specifically, our algorithms for energy clustering perform 
closely to $k$-means and GMM on normally distributed data, but on the other
hand, performs considerably better than $k$-means and GMM on  data that 
is not normally distributed. It also performs better than $k$-means 
and GMM in high dimensions, even on Gaussian settings. However, it performs 
worse than GMM, and more closely to $k$-means, for highly unbalanced clusters
with normal distributions.

%Our work is organized as follows. In section~\ref{sec:background}, we introduce
%the necessary background on energy statistics and RKHS.
%Section~\ref{sec:clustering_theory} contains the main theoretical 
%results of this paper,
%where we consider a clustering theory based on energy statistics leading
%to a QCQP, which is NP-hard in general. We also show the equivalence to 
%kernel $k$-means optimization problem.
%In Section~\ref{sec:weighted}, we generalize these results to a weighted
%version of energy statistics, providing connections to graph
%partitioning problems.
%In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
%problem in one dimension,
%where we propose an algorithm which requires no random initialization.
%In section~\ref{sec:algo}, we consider Lloyd's and Hartigan's methods
%and propose iterative algorithms to solve the QCQP.
%Section~\ref{sec:numerics} contains some carefully designed numerical
%experiments showing that, in a variety of settings, energy clustering based
%on Hartigan's algorithm
%outperform $k$-means, GMM, kernel $k$-means, and a version of
%spectral clustering.
%Our final conclusions are presented in Section~\ref{sec:conclusion}.

@@

The energy distance can be even further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the associated Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a proper metric.

@@

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Due to \eqref{eq:mmd} we can compute the energy 
distance $\mathcal{E}(P, Q)$ between two probability distributions
using the inner 
product of $\Hk$.

Finally, let us recall the main formulas from energy statistics
for the test statistic of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ where
$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y) ,
\end{equation}
where $n_i = |\C_i|$ is the number of elements in partition
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a test statistic for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
Note that this test does not make any assumptions
about the distributions $P_j$, thus it is said to be non-parametric or
distribution-free.

@@

% @gui: maybe common in math, but certainly not in stats / data science. 
% in any case, what is the package that does that?
% @jovo: \usepackage{mathtools} \mathtoolsset{showonlyrefs}

% @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
% @jovo: I'll put a comment below, I simply removed the previous phrase,
% didn't say anything instead.

@@

Due to the energy test statistic for equality of distributions,
the obvious
criterion for clustering data is to 
maximize $S$ which makes 
each cluster as different
as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, the test statistic $S$ should attain a maximum when 
each point is correctly
classified as belonging to the cluster associated to its probability
distribution.
The following 
straightforward result
shows that maximizing $S$ is, however, equivalent to minimizing
$W$ which has a more convenient form.

@@

Based on Proposition~\ref{th:qcqp2}, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
This optimization problem is nonconvex, and also NP-hard,
thus a direct approach 
is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints, or obtaining a relaxed SDP version of it.
For instance, the relaxed problem
\begin{equation}
\label{eq:relaxed}
\max_{Y} \Tr \left( Y^\top G \, Y \right) \quad \mbox{s.t. $Y^\top Y = I$}
\end{equation}
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U \in \mathbb{R}^{n\times k}$ 
contain the top $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting
optimal objective function assumes the value 
$\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
\sum_{i=1}^k \lambda_i$.

Spectral clustering is based on the above approach, where
one further normalize the rows of $Y^\star$, 
i.e. $Y^\star \to \widetilde{Y}$ where 
$\widetilde{Y}_{ij} \equiv Y^\star_{ij} \big/ \| Y^\star_{i\bullet} \|$. 
The vectors
$y_i \equiv \widetilde{Y}_{i\bullet} \in \mathbb{R}^k$, which are the
rows of matrix $\widetilde{Y}$, lie on the surface of the $k$-dimensional 
unit sphere.
The set $\{ y_i \}_{i=1}^n$ is a spectral embedding of the original
set $\{ x_i \}_{i=1}^n$ onto an eigen-subspace of the kernel matrix $G$.
After performing this embedding, one group the points 
$\{ y_i \}_{i=1}^n$ into $k$ clusters 
$\{ \tC_1,\dotsc,\tC_k \}$ using any clustering algorithm, such as
$k$-means. This yields the final solution
where the original point $x_i$ belongs
to cluster $\C_j$ if and only if $y_i$ belongs to $\tC_j$.
A procedure on these lines was proposed in the seminal paper
\cite{NgJordan}, which considers the eigen-decomposition of
the matrix $D^{-1/2} G D^{-1/2}$ instead of $G$, and also 
where $G$ is an affinity matrix with the modification that
its diagonal elements are set to zero $G_{ii}=0$, 
and $D=\diag(d_{1},\dotsc,d_{n})$ is the degree matrix, 
with $d_{i} \equiv \sum_{j=1}^n G_{ij}$. 
We consider a slightly modification of this procedure,
which we call \mbox{$\mathcal{E}^{S}$-clustering} 
to emphasize that its a spectral
method whose kernel comes from energy statistics, by 
computing the 
eigen-decomposition of $G D^{-1}$ instead.
This  method is described in 
Algorithm~\ref{spectralalgo} and will be used as a comparison
to the other iterative algorithms 
that will be proposed later.

\begin{figure}
\begin{flushleft}
\begin{algorithm}[H]
\vspace{.5em}
\begin{algorithmic}[1]
\INPUT Gram matrix $G\in\mathbb{R}^{n\times n}$, and number of clusters $k$
\OUTPUT label matrix $Z \in \{ 0, 1\}^{n \times k}$
\STATE set the diagonal of $G$ to zero: $G_{ii} \leftarrow 0$ 
        for $i=1,\dotsc,n$
\STATE $D\leftarrow\diag(d_1,\dotsc,d_n)$ 
        where $d_i \leftarrow \sum_{j=1}^n G_{ij}$
\STATE solve the generalized eigenvalue problem $G v_i = \lambda_i D v_i$,
        where $i=1,\dotsc,n$ with eigenvalues, and corresponding eigenvectors,
        obeying the ordering
        $\lambda_1 \ge \lambda_2 \ge \dotsc \lambda_n$
\STATE $Y \leftarrow (D v_1| \dotsm | D v_k) 
        \in \mathbb{R}^{n\times k}$ 
has the top $k$ eigenvectors (multiplied by $D$) as columns
\STATE normalize the rows $Y$, i.e. 
$\widetilde{Y}_{ij} \leftarrow Y_{ij} \Big/ 
\left(\sum_{j=1}^n |Y_{ij}|^2\right)^{1/2}$
\STATE apply $k$-means on $\{ y_i \}_{i=1}^n$, 
where $y_i = \widetilde{Y}_{i\bullet}$
is the $i$th row of $\widetilde{Y}$
\STATE $Z_{ij} \leftarrow 1$ if $y_i$ is in cluster $j$, 
        and $Z_{ij} \leftarrow 0$ otherwise
\end{algorithmic}
\caption{
\label{spectralalgo}
$\mathcal{E^S}$-clustering is an exact spectral method to solve
the relaxed problem \eqref{eq:relaxed}.
\hspace{\fill}
}
\end{algorithm}
\end{flushleft}
\end{figure}

Note that the energy clustering problem \eqref{eq:qcqp2} 
is valid for data living in an \emph{arbitrary} space of negative type, where
a semimetric $\rho$, and thus the kernel $\kk$, are
assumed to be known. Standard energy statistics in
Euclidean space fixes a family of choices 
$\rho_\alpha(x,y) = \| x - y\|^\alpha$, with the corresponding
kernel $K_\alpha(x,y) = \tfrac{1}{2} \left( \| x \|^\alpha + \| y \|^\alpha - 
\| x-y\|^\alpha \right)$, where we
set $x_0=0$ in \eqref{eq:kernel_semimetric} and  $0<\alpha\le 2$.
The same would be valid for data living in a more general
metric space $(\mathcal{X}, \rho)$, where the corresponding semimetric $\rho$ 
fixes the kernel.
In any case, the energy  clustering 
formulation does not make assumptions
about  the distribution of the data, 
contrary to $k$-means and GMM, for example.
In practice, however,
the clustering quality strongly depend on the choice of a suitable
$\rho$, which measures the similarity between data points,
and is equivalent to choosing an appropriate kernel.
However, if prior information is available to choose an appropriate $\rho$,
it can be conveniently incorporated in the 
optimization problem \eqref{eq:qcqp2}.

@@

In this section we introduce an iterative algorithm to find a local
maximizer of the optimization problem \eqref{eq:qcqp2}. Due to 
Proposition~\ref{th:kernel_kmeans} we can also find an approximate
solution by the well-known kernel $k$-means algorithm based
on Lloyd's heuristic (see \cite{Dhillon2,Dhillon}), which 
for convenience will also be restated in the present context.

>>>>

\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
%\usepackage{nicefrac}       % compact symbols for 1/2, etc.
%\usepackage{microtype}      % microtypography
%\usepackage{mathtools}
%\usepackage{printlen}
%\usepackage{cite}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs}
\usepackage[inline]{enumitem}
\usepackage[inline]{enumitem}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}
\usepackage{algorithmic}
\usepackage{algorithm}
%\usepackage[none]{hyphenat}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array,booktabs}
\usepackage{xcolor}
\usepackage{makecell}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

@@

\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by the 
Newtonian gravitational potential from classical mechanics, and it provides 
a 
hypothesis test for equality of distributions. 
It was further generalized from Euclidean spaces to metric spaces of 
strong negative type, and more recently, a connection with reproducing 
kernel Hilbert spaces (RKHS) was established. 
Here we consider the clustering problem from an 
energy statistics theory perspective, 
providing a precise mathematical formulation 
yielding a quadratically constrained 
quadratic program (QCQP) in the associated RKHS, thus establishing the
connection with kernel methods. We show that this QCQP
is equivalent to kernel $k$-means optimization problem once the kernel
is fixed.
These results imply a first principles derivation of kernel $k$-means 
from energy statistics.
However, energy statistics fixes a family of standard kernels.
Furthermore, we also consider a weighted version of energy statistics, 
making connection to graph partitioning problems.
To find local optimizers of such QCQP we propose an iterative algorithm based 
on Hartigan's method, which in this case has the same computational cost 
as kernel $k$-means algorithm, based on Lloyd's heuristic, but usually 
with better clustering quality. 
We provide carefully designed numerical experiments showing the superiority 
of the proposed method compared to kernel $k$-means, spectral clustering,
standard $k$-means, and Gaussian mixture models in a variety of settings.
\end{abstract}

@@

Energy statistics \cite{Szkely2013}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational potential in classical mechanics.
It provides a model-free hypothesis test for equality of 
distributions which is achieved 
under minimum energy. When probability distributions are different the 
statistical potential energy diverges as sample size increases, while tends 
to a nondegenerate limit distribution when probability
distributions are equal. 
Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering \cite{RizzoClustering} 
by extending Ward's method of minimum variance. Moreover, 
in Euclidean spaces, an application of 
energy statistics to clustering
was already proposed \cite{Kgroups}. 
We refer the reader to \cite{Szkely2013}, 
and references therein, for an overview of energy statistics theory and 
its applications.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
The notion of distance covariance was further generalized from Euclidean 
spaces to metric spaces of strong negative type \cite{Lyons}. Furthermore, 
the missing link between energy distance based tests and kernel 
based tests has 
been recently resolved \cite{Sejdinovic2013}, 
establishing an equivalence between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS).
This equivalence immediately relates energy statistics to
kernel methods often used in machine learning, and form the basis 
of our approach.

@@

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen. A well-known kernel based clustering method
is kernel $k$-means, which is preciselly $k$-means 
formulated in the feature space \cite{Girolami}. 
Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd}
of grouping points that are closer to a cluster center.
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics, in Euclidean spaces,
was considered in \cite{Kgroups}, 
the precise optimization problem behind this approach
remains elusive, as well as the connection with kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. This immediatelly
establishes the connection between energy statistics based clustering,
or \emph{energy clustering} for short, with kernel methods. Moreover,
our formulation holds for general semimetric spaces of negative type.
We also show that such QCQP is equivalent to kernel $k$-means 
optimization problem, however, the kernel is fixed by 
energy statistics.
The equivalence between kernel $k$-means, spectral clustering, and graph 
partitioning problems is well-known \cite{Dhillon,Dhillon2}. We further
demonstrate 
how these relations arise from a weighted version of energy statistics.

Our main 
algorithmic contribution is to use Hartigan's method \cite{Hartigan} to 
find local solutions of the above mentioned QCQP, which is NP-hard in general.
Hartigan's method was also used in \cite{Kgroups}, but without
any connection to kernels. More importantly, the
advantages
of Hartigan's over Lloyd's method was already demonstrated 
in some simple settings
\cite{Telgarsky,Slonin}, but apparently this method did not receive 
the deserved attention. To the 
best of our knowledge, Hartigan's method was not previously 
employed together with kernel methods. 
We provide a fully kernel based Hartigan's algorithm for clustering,
where the kernel is fixed by energy statistics. 
We make clear the advantages of this proposal versus 
Lloyd's method, which kernel $k$-means is based upon and will also be used 
to solve our QCQP. We show that both algorithms  have the same
time complexity, but Hartigan's method in kernel spaces offer several
advantages. Furthermore, in the examples considered in this paper, it 
also provides superior performance compared spectral clustering, which
is more expensive and in fact solves a relaxed version of our QCQP.

Our numerical results provide compelling evidence that 
Hartigan's method applied to energy clustering 
is more accurate and robust than kernel $k$-means algorithm.
Furthermore, our experiments illustrate 
the flexibility of energy clustering, 
showing that it is able to perform accurately on data coming from 
very different distributions, contrary to $k$-means and GMM for instance.
More specifically, the proposed method performs 
closely to $k$-means and GMM on normally distributed data, however,
it is significantly better on data that 
is not normally distributed. 
Its superiority in high dimensions
is striking, being more accurate than $k$-means and GMM 
even on Gaussian settings.

@@

The energy distance can be even further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the associated Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a proper metric. 
Our energy clustering formulation, proposed in the next section,
will be based on the generalized
energy distance \eqref{eq:energy3}.

@@

Now we can state the equivalence between the generalized 
energy distance \eqref{eq:energy3} and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Due to \eqref{eq:mmd} we can compute the energy 
distance $\mathcal{E}(P, Q)$ between two probability distributions
using the inner 
product of $\Hk$.

Finally, let us recall the main formulas from generalized energy statistics
for the test statistic of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ where
$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y) ,
\end{equation}
where $n_i = |\C_i|$ is the number of elements in partition
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a test statistic for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
Note that this test does not make any assumptions
about the distributions $P_j$, thus it is said to be non-parametric or
distribution-free.

@@

@@

Due to the previous test statistic for equality of distributions,
the obvious
criterion for clustering data is to 
maximize $S$ which makes 
each cluster as different
as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, the test statistic $S$ should attain a maximum when 
each point is correctly
classified as belonging to the cluster associated to its probability
distribution.
The following 
straightforward result
shows that maximizing $S$ is, however, equivalent to minimizing
$W$ which has a more convenient form.

@@

Based on Proposition~\ref{th:qcqp2}, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
This optimization problem is nonconvex, and also NP-hard,
thus a direct approach 
is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints, or obtaining a relaxed SDP version of it.
For instance, the relaxed problem
\begin{equation}
\label{eq:relaxed}
\max_{Y} \Tr \left( Y^\top G \, Y \right) \qquad \mbox{s.t. $Y^\top Y = I$}
\end{equation}
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U \in \mathbb{R}^{n\times k}$ 
contain the top $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting
optimal objective function assumes the value 
$\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
\sum_{i=1}^k \lambda_i$. 
Spectral clustering is based on the above approach, where
one further normalize the rows of $Y^\star$, then cluster
the resulting rows as data points.
A procedure on these lines was proposed in the seminal papers
\cite{Malik,NgJordan}.

Note that the optimization problem \eqref{eq:qcqp2} based on energy
statistics
is valid for data living in an \emph{arbitrary} space of negative type, where
a semimetric $\rho$, and thus the kernel $\kk$, are
assumed to be known. The standard energy distance \eqref{eq:energy2}
fixes a family of choices in Euclidean spaces given by
\begin{equation}
\rho_\alpha(x,y) = \| x - y\|^\alpha, \qquad 
K_\alpha(x,y) = \tfrac{1}{2} \left( \| x \|^\alpha + \| y \|^\alpha - 
\| x-y\|^\alpha \right) ,
\end{equation}
for $0<\alpha\le 2$ and 
we fix $x_0=0$ in \eqref{eq:kernel_semimetric}.
The same would be valid for data living in a more general
semimetric space $(\mathcal{X}, \rho)$ where $\rho$ 
fixes the kernel.
In practice, the clustering quality strongly depend on the choice 
of a suitable $\rho$. 
Nevertheless, if prior information is available to make this choice, it
can be immediately incorporated in the 
optimization problem \eqref{eq:qcqp2}.

@@

In this section we introduce an iterative algorithm to find a local
maximizer of the optimization problem \eqref{eq:qcqp2}. Due to 
Proposition~\ref{th:kernel_kmeans} we can also find an approximate
solution by the well-known kernel $k$-means algorithm based
on Lloyd's heuristic \cite{Dhillon2,Dhillon}, which 
for convenience will also be restated in the present context.
